Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-08-12 analysis/paper/paper.qmd
updates_figures

p_raw_material <- 
ggplot(df_long %>% 
         mutate(category = case_when(
           category == "cores" ~ "Core",
           category == "spheroids" ~ "Spheroid",
           category == "total" ~ "Total",
           category == "Polys_and_Multi_cores" ~ "Polys and Multi cores"
           )), 
       aes(x = material, 
           y = prop, 
           fill = category)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(
    x = "",
    y = "Percentage (%)",
    fill = "Technological Type"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
    scale_fill_manual(values = c(
    "Core" = "#66c2a5", 
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "lightgray",
    "Total" = "gray"
  )) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = c(0.95, 0.95),              
    legend.justification = c("right", "top"),      
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
    legend.box.background = element_rect(color = "black")  
  )
p_raw_material

@@

long_data <- data %>%
  select(core_length, core_width, core_thickness,
         spheroid_length, spheroid_width, spheroid_thickness,
         multi_poly_length, multi_poly_width, multi_poly_thickness) %>%
  pivot_longer(
    everything(),
    names_to = "type_metric",
    values_to = "value"
  ) %>%
  extract(type_metric, into = c("artifact", "metric"),
          regex = "^(core|spheroid|multi_poly)_(length|width|thickness)") %>%
  filter(!is.na(value)) %>%
  mutate(
    artifact = recode(artifact,
                      "core" = "Core",
                      "spheroid" = "Spheroid",
                      "multi_poly" = "Polys and Multi cores"),
    metric = factor(metric, levels = c("length", "width", "thickness"))
  )

# 画图
p_dimensions <- 
ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.5) +
  labs(
    x = "",
    y = "Measurement (mm)",
    fill = "Technological Type"
  ) +
  scale_fill_manual(values = c(
    "Core" = "#66c2a5", 
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "lightgray"
  )) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = c(0.95, 0.95),       
    legend.justification = c("right", "top"),  
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA), 
    legend.box.background = element_rect(color = "black")  
  )

In terms of dimensions, the average length, width, thickness, and mass of spheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-basic-characteristics. Two spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions. In contrast, cores displayed greater variability in size, with standard deviations of `r round(summary_stats$core_length_sd, 1)` mm, `r round(summary_stats$core_width_sd, 1)` mm, and `r round(summary_stats$core_thickness_sd, 1)` mm for length, width, and thickness, respectively. The polyhedrons and multifacial cores showed less variation than the overall core group, but their dimensional variability remained substantially greater than that of the spheroids. Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry. In contrast, cores, including polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.

@@

From the normalized spherical harmonic coefficients, we computed the rotation-invariant power spectrum by summing the squared magnitudes of the coefficients at each degree. Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor. To explore variation between models, we applied UMAP for dimensionality reduction and visualization [@Van2023]. We included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8) and box (1.2:1:1), and disc-like (flattened, and like discoid core)—in the SPHARM analysis and UMAP as controls to aid in the interpretation of the results. Spheroids and cores were positioned in this morphospace based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.

@@

### Centroid offset

We also used the distance between its centre of mass (CoM) and the centre of its bounding box (CoBB) as a measure of volumetric symmetry [@Muller2023]. A perfect sphere will have its CoM and CoBB perfectly aligned, resulting in a zero offset. As the object becomes more irregular or asymmetric, this offset increases. For each 3D model, the CoBB is defined as the midpoint of the maximum orthogonal dimensions—length (y), width (x), and thickness (z)—of its minimum bounding box. The CoM, which approximates the true mass center for objects made of homogeneous materials, is computed by weighting each triangular face of the mesh by its surface area and normalizing the result. The distance between CoM and CoBB, normalized by the artifact’s volume, serves as a quantitative indicator of sphericity: lower values indicate more spherical forms. <!-- BM: delete all mentions of Centroid offset -->

@@

# 画图
ggplot(data_clean, aes(x = Sphericity, y = `SPHARM power`)) +
  # 散点（深红色）
  geom_point(size = 3, alpha = 0.9, color = "#B22222") +

  # 拟合线（深蓝灰）
  geom_smooth(method = "lm", se = TRUE, color = "#1C3041", fill = "#A3B1C6", linetype = "solid") +

   scale_x_continuous(breaks = seq(0.4, 1.0, by = 0.04)) +

  # 自定义配色主题
  theme_minimal(base_size = 15) +
  theme(
    panel.grid.major = element_line(color = "#DDDDDD"),
    panel.grid.minor = element_blank(),
    axis.title = element_text(color = "#333333"),
    axis.text = element_text(color = "#333333")
  ) +

  # 标题与坐标轴标签
  labs(
    x = "Sphericity",
    y = "SPHARM Power"
  ) +
  annotate("text",
         x = Inf, y = Inf,
         hjust = 1.1, vjust = 1.2,
         label = paste0("r = ", round(correlation, 3),
                        "\n", "p = ", pretty_print_sci(p_value),
                        "\n", "n = ", nrow(data_clean)),
         size = 5, color = "#1C3041")
```

@@

```{r, echo=FALSE}
library(here)
# 读取无列名的csv
library(here)
data_dir <- here("analysis", "data", "derived_data", "power_spectrum")
csv_files <- list.files(data_dir, pattern = "\\.csv$", full.names = TRUE)

@@

@@

# 绘图，带误差带
ggplot(summary_df_filtered, aes(x = degree, y = mean_power, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(
    ymin = pmax(mean_power - sd_power, 1e-10),
    ymax = mean_power + sd_power
  ), alpha = 0.2, linetype = 0) +
  scale_y_log10(
  breaks = c(1e-8, 1e-6, 1e-4, 1e-2, 1),
  labels = parse(text = c("10^-8", "10^-6", "10^-4", "10^-2", "10^0"))
  )+
  scale_x_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 4),
    expand = c(0, 0)
  ) +
  coord_cartesian(clip = "on") +
  scale_color_manual(values = c(
    "All specimens" = "#66c2a5",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "lightgray"
  )) +
  scale_fill_manual(values = c(
    "All specimens" = "#66c2a5",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "lightgray"
  )) +
  labs(
    title = "Mean Power Spectrum with Standard Deviation",
    x = "Degree (l)",
    y = "Mean Total Power (log scale)",
    color = "Group",
    fill = "Group"
  ) +
  theme_minimal() +
  theme(plot.margin = margin(0, 0, 0, 0))
```

We calculated the spherical sphericity index, harmonics analysis, and centroid offset measurements to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].

Our sphericity analysis reveals a clear difference between spheroids and multifacial cores/polyhedra. The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of multifacial cores and polyhedron (`r mean_sphericity_poly_multi`), with a Mann-Whitney U test yielding U = `r U_value`, p = $`r p_value_Mann_sci`$. This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification as a distinct type.

@@

We performed a Variance Analysis on the power of each harmonic degree. The results indicate that l=2 and l=3 exhibit the highest variance across the sample, suggesting that these low-order terms capture the most significant shape variation. Specifically, l=2 reflects global asymmetries—such as elongation or flattening—while degree 3 is more sensitive to localized features like curvature and edge protrusions. <!-- BM: can we show the results of this somehow, e.g. with a figure? -->

@@

The UMAP-based groupings also correspond to significant differences in sphericity, as indicated by a Kruskal–Wallis test ($\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$). The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation (@fig-Correlation-SPHARMpower-Sphericity).

The spherical harmonic energy (SHE), a novel metric introduced here, quantifies the geometric approximation of core morphology to an ideal sphere. Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity. Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy of sphericity. <!-- BM: consider a line plot to show all artefacts, one line per artefact -->

>>>>

p_raw_material <- 
ggplot(df_long %>% 
         mutate(category = case_when(
           category == "cores" ~ "Cores",
           category == "spheroids" ~ "Spheroid",
           category == "total" ~ "Total",
           category == "Polys_and_Multi_cores" ~ "Polys and Multi cores"
           )), 
       aes(x = material, 
           y = prop, 
           fill = category)) +
  geom_col(position = position_dodge(width = 0.6), width = 0.5) +
  labs(
    x = "",
    y = "Percentage (%)",
    fill = "Technological Type"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
    scale_fill_manual(values = c(
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "#66c2a5",
    "Cores" = "lightgray", 
    "Total" = "gray"
  )) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = c(0.95, 0.95),              
    legend.justification = c("right", "top"),      
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA),  
    legend.box.background = element_rect(color = "black")  
  )
p_raw_material

@@

long_data <- data %>%
  select(core_length, core_width, core_thickness,
         spheroid_length, spheroid_width, spheroid_thickness,
         multi_poly_length, multi_poly_width, multi_poly_thickness) %>%
  pivot_longer(
    everything(),
    names_to = "type_metric",
    values_to = "value"
  ) %>%
  extract(type_metric, into = c("artifact", "metric"),
          regex = "^(core|spheroid|multi_poly)_(length|width|thickness)") %>%
  filter(!is.na(value)) %>%
  mutate(
    artifact = recode(artifact,
                      "core" = "Cores",
                      "spheroid" = "Spheroid",
                      "multi_poly" = "Polys and Multi cores"),
    metric = factor(metric, levels = c("length", "width", "thickness"))
  )

# 画图
p_dimensions <- 
ggplot(long_data, aes(x = metric, y = value, fill = artifact)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.5) +
  labs(
    x = "",
    y = "Measurement (mm)",
    fill = "Technological Type"
  ) +
  scale_fill_manual(values = c(
    "Cores" = "lightgray", 
    "Spheroid" = "orange", 
    "Polys and Multi cores" = "#66c2a5"
  )) +
  theme_minimal(base_size = 8) +
  theme(
    legend.position = c(0.95, 0.95),       
    legend.justification = c("right", "top"),  
    legend.background = element_rect(fill = alpha("white", 0.6), color = NA), 
    legend.box.background = element_rect(color = "black")  
  )

p_dimensions
```

In terms of dimensions, the average length, width, thickness, and mass of spheroids were `r round(summary_stats$spheroid_length_mean, 1)` mm (SD = `r round(summary_stats$spheroid_length_sd, 1)`), `r round(summary_stats$spheroid_width_mean, 1)` mm (SD = `r round(summary_stats$spheroid_width_sd, 1)`), `r round(summary_stats$spheroid_thickness_mean, 1)` mm (SD = `r round(summary_stats$spheroid_thickness_sd, 1)`), and `r round(summary_stats$spheroid_weight_mean, 1)` g (SD = `r round(summary_stats$spheroid_weight_sd, 1)`), respectively, as shown in @fig-basic-characteristics. Two spheroids showed outlier values in size, being either unusually large or small, while the rest exhibited relatively concentrated size distributions. In contrast, polyhedrons and multifacial cores displayed greater variability in size, with standard deviations of `r round(summary_stats$multi_poly_length_sd, 1)` mm, `r round(summary_stats$multi_poly_width_sd, 1)` mm, and `r round(summary_stats$multi_poly_thickness_sd, 1)` mm for length, width, and thickness, respectively. Additionally, spheroids tend to have similar values across length, width, and thickness, reflecting their overall symmetry. In contrast, polyhedrons and multifacial cores show marked differences among these dimensions, indicating more irregular and elongated forms.

@@

From the normalized spherical harmonic coefficients, we computed the rotation-invariant power spectrum by summing the squared magnitudes of the coefficients at each degree. Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor. To explore relationship between models, we applied Uniform Manifold Approximation and Projection (UMAP) for dimensionality reduction and visualization [@Van2023]. Unlike traditional linear dimensionality reduction methods such as Principal Component Analysis (PCA), which capture only global linear variance patterns, UMAP is a non-linear technique designed to preserve both local and global data structure. Its grouping is based on the local and global similarity of shape data. This makes UMAP particularly well suited for complex shape data represented by spherical harmonic coefficients, where morphological variations may manifest in subtle, non-linear ways across multiple degrees. By effectively capturing these non-linear relationships, UMAP yields a refined morphospace representation, enabling detection of complex shape differences that linear techniques may overlook.

To better interpret the grouping revealed by UMAP, we included a set of standard geometric models—such as sphere, rounded cube, ellipsoid (1.2:1:0.8), box (1.2:1:1), and disc-like forms (flattened and discoid core)—in the SPHARM analysis and UMAP embedding as visual references. Spheroids and cores were positioned in this morphospace based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.

@@

@@

ggplot(data_clean, aes(x = Sphericity, y = `SPHARM power`)) +
  geom_point(size = 3, alpha = 0.9, color = "#F28E2B") +
  geom_smooth(method = "lm", se = TRUE, color = "#1C3041", fill = "#A6CEE3") +
  scale_x_continuous(breaks = seq(0.4, 1.0, by = 0.04)) +
  theme_minimal(base_size = 15) +
  theme(
    panel.grid.major = element_line(color = "#DDDDDD"),
    panel.grid.minor = element_blank(),
    axis.title = element_text(color = "#333333"),
    axis.text = element_text(color = "#333333")
  ) +
  labs(x = "Sphericity", y = "SHE") +
  annotate("text",
         x = Inf, y = Inf,
         hjust = 1.1, vjust = 1.2,
         label = paste0("r = ", round(correlation, 3),
                        "\n", "p = ", pretty_print_sci(p_value),
                        "\n", "n = ", nrow(data_clean)),
         size = 5, color = "#1C3041")
```

```{r, include=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)

# 把两个指标转成长格式，方便facet_wrap显示
df_long <- mean_sphericity_tbl %>%
  select(Group, Sphericity, `SPHARM power`) %>%
  pivot_longer(cols = c(Sphericity, `SPHARM power`),
               names_to = "Variable",
               values_to = "Value")

# 自定义Group名称映射，方便图例显示
df_long <- df_long %>%
  mutate(Group = factor(Group, levels = c("Spheroid + Subspheroid", "Polyhedron + Multifacial"),
                        labels = c("Spheroid + Subspheroid", "Polyhedron + Multifacial")))

# 画图
ggplot(df_long, aes(x = Group, y = Value, fill = Group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 1, size = 1) +
  facet_wrap(~ Variable, scales = "free_y") +
  scale_fill_manual(values = c("Spheroid + Subspheroid" = "orange", "Polyhedron + Multifacial" = "#66c2a5")) +
  labs(x = "", y = "Value", fill = "Group", title = "Sphericity and SHE by Group") +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )
```

```{r, include=FALSE}
# 读取 CSV
df <- read_csv(
  here("analysis", "data", "derived_data", "variance_per_degree.csv")
)

ggplot(df, aes(x = degree, y = variance)) +
  geom_line(color = "#1f78b4", size = 1.3, alpha = 0.8) +          
  geom_point(color = "#ff7f00", size = 3, alpha = 0.9, shape = 21, fill = "#ffbb78", stroke = 1) +  # 橙色边框，填充浅橙
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +    
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8)) +      
  labs(
    title = "Variance per Degree",
    x = "Spherical Harmonic Degree",
    y = "Variance"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "#DDDDDD", size = 0.5),
    panel.grid.minor = element_blank()
  )

@@

library(here)
data_dir <- here("analysis", "data", "derived_data", "power_spectrum")
csv_files <- list.files(data_dir, pattern = "\\.csv$", full.names = TRUE)

@@

#l_2_3_power <- all_data %>%
  #filter(degree %in% c(2, 3)) %>%
  #select(file, degree, total_power) %>%
  #pivot_wider(names_from = degree, values_from = total_power, names_prefix = "l_")

# 查看结果
#View(l_2_3_power)

@@

# 绘图，带误差带
ggplot(summary_df, aes(x = degree, y = mean_power, color = group, fill = group)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(
    ymin = pmax(mean_power - sd_power, 1e-10),
    ymax = mean_power + sd_power
  ), alpha = 0.2, linetype = 0) +
  scale_y_log10(
  breaks = c(1e-8, 1e-6, 1e-4, 1e-2, 1),
  labels = parse(text = c("10^-8", "10^-6", "10^-4", "10^-2", "10^0"))
  )+
  scale_x_continuous(
    limits = c(0, 20),
    breaks = seq(0, 20, by = 4),
    expand = c(0, 0)
  ) +
  coord_cartesian(clip = "on") +
  scale_color_manual(values = c(
    "All specimens" = "lightgray",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "#66c2a5"
  )) +
  scale_fill_manual(values = c(
    "All specimens" = "lightgray",
    "Spheroid/Subspheroid" = "orange",
    "Polyhedron/Multifacial" = "#66c2a5"
  )) +
  labs(
    title = "Mean Power Spectrum with Standard Deviation",
    x = "Degree (l)",
    y = "Mean Total Power (log scale)",
    color = "Group",
    fill = "Group"
  ) +
  theme_minimal() +
  theme(plot.margin = margin(t = 10, r = 10, b = 10, l = 10),
  legend.position = c(0.95, 0.95),
  legend.justification = c("right", "top"),
  legend.background = element_rect(fill = alpha("white", 0.8), color = NA),
  legend.box.background = element_rect(color = "gray80"),
  legend.box.margin = margin(5, 5, 5, 5)
  )
```

```{r, include=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)

# 读取 Excel 文件
file_path <- "E:/Lithic Analysis/QSYSpheroidsStudy/analysis/data/derived_data/umap_l2_l3.XLSX"
df <- read_excel(file_path)

# 查看数据结构
print(head(df))

# 计算相关系数矩阵
correlations <- df %>%
  select(UMPA1, UMAP2, degree1, degree2) %>%
  cor()

print("相关系数矩阵：")
print(correlations)

# 将数据长格式化方便绘图
df_long <- df %>%
  pivot_longer(cols = c(degree1, degree2), names_to = "degree", values_to = "degree_value")

# 绘制 UMAP1 和 degree的散点图
p1 <- ggplot(df_long, aes(x = UMPA1, y = degree_value, color = degree)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "UMAP1 vs degree1 (l=2) and degree2 (l=3)",
       x = "UMAP1",
       y = "Degree Power") +
  theme_minimal()

# 绘制 UMAP2 和 degree的散点图
p2 <- ggplot(df_long, aes(x = UMAP2, y = degree_value, color = degree)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "UMAP2 vs degree1 (l=2) and degree2 (l=3)",
       x = "UMAP2",
       y = "Degree Power") +
  theme_minimal()

print(p1)
print(p2)

```
We calculated the spherical sphericity ans spherical harmonic energy (SHE) to both the spheroids (n = `r n_total_typology`) and all other multifacial and polyhedron cores (n = `r n_poly_multi`) in the Qianshangying assemblage. These analyses aim to assess whether spheroids are morphologically distinct from other cores, and whether they exhibit standardized shapes—an essential criterion for evaluating the existence of deliberate shaping strategies and a goal product. In addition, we compare quantitative outcomes with traditional typological classifications to assess their consistency. We present a summary table listing each core's blank morphology and type, final morphology, and cortex coverage here in [@tbl-1].

Our results of sphericity reveals a clear difference between spheroids and multifacial cores/polyhedra. The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of multifacial cores and polyhedron (`r mean_sphericity_poly_multi`), with a Mann-Whitney U test yielding U = `r U_value`, p = $`r p_value_Mann_sci`$. This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification as a distinct type.

@@

Subsequently, we computed the rotation-invariant power spectrum. Here we present the power spectrum of different groups, including spheroids, multifacial cores, polyhedrons, and the total of all these types combined. The results show that the spheroid group has significantly lower power than polyhedrons and multifacial cores at both low and high degrees, indicating that spheroids possess a more symmetrical overall shape and more regular local features. In addition, the spheroid group exhibits smaller inter-group morphological variation than polyhedrons and multifacial cores, particularly at low degrees. Then we performed a Variance Analysis on the power of each harmonic degree. The results indicate that l=2 and l=3 exhibit the highest variance across the sample, suggesting that these low-order terms capture the most significant shape variation. Specifically, l=2 reflects global asymmetries, such as elongation or flattening, while degree 3 is more sensitive to localized features like curvature and edge protrusions.

@@

The UMAP-based groupings also correspond to significant differences in sphericity, as indicated by a Kruskal–Wallis test ($\chi^2$ = `r round(H_value, 2)`, df = `r df_kruskal`, p = $`r p_value_kruskal_sci`$). The result shows statistically significant differences in sphericity among the shape clusters, further supporting the effectiveness of the spherical harmonic approach in capturing meaningful morphological variation.

The spherical harmonic energy (SHE), a novel metric introduced here, quantifies the geometric approximation of core morphology to an ideal sphere. Normalized SHE approaching unity indicates progressively spherical configurations, while deviations from unity reflect morphological irregularity. Our analysis demonstrates significant negative correlation between SHE and classic sphericity measures (r = `r round(correlation, 2)`, df = `r df_value`, p = $`r p_value_pearson_sci`$)), validating its efficacy as a proxy of sphericity (@fig-Correlation-SPHARMpower-Sphericity).
